---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data
The data includes different types of crime in Prince George's County, such as accidents, assaults, stolen vehicles, homicide, robbery, and sex offenses. Some categories are more important from a reporting standpoint because serious crimes like homicide and assault can have a major impact on the community. The different categories can also make summarizing the data more difficult because there are many different types of crimes to compare.

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?



### Task 4: Calculating Averages

Add up the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime using the total number of months, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.


```{r}

pg_crime_averages <- pg_crime |> 
  mutate(
    total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec,
    average_incidents = total_incidents / 12
  )
### Task 5: Describe the Averages

The average number of crimes is different for each category. Some crimes happen more often than others, while serious crimes like homicide happen less often. This shows that crime in Prince George’s County varies by category.


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> 
  mutate(
    june_change = (June - average_incidents) / average_incidents * 100,
    july_change = (July - average_incidents) / average_incidents * 100,
    august_change = (August - average_incidents) / average_incidents * 100
  )
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

ANSWER HERE

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

The data shows the number of reported incidents for each crime category across the 12 months of 2024. Looking at the monthly totals and averages helps show which types of crime occurred most often and how crime varied throughout the year.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
