---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r} 
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv") 
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The data contains information about White House staff members, including their names, status, salaries, pay basis and position titles. The salary amounts stand out because there is a large difference between what some staff members earn. There are also some salaries listed as $0, which could cause problems when analyzing the data

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> 
  summarize(
    mean_salary = mean(SALARY),
    median_salary = median(SALARY)
  )


```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries|>
  summarize(
    mean_salary=mean(SALARY),
    median_salary= median(SALARY),
    min_salary= min(SALARY),
    max_salary= max(SALARY),
    
  )
  
   
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean salary is $112,702.70, while the median salary is $109,243. This shows that the typical salary is around $109,000 to $113,000. The lowest salary is $0 and the highest salary is $197,200, showing a large difference between the lowest and highest salaries. The $0 salaries and the very high salaries could be outliers in the data.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

My story idea is to look at why some White House workers make more money than others. My next step would be to look at the highest and lowest-paid workers and their jobs.
